Learning to Estimate Search Progress Using Graph Neural Networks \thanks{This work expands on \citet{sudry2022learning}.}

### 2024-11-05 report/submission.tex
pushing new report with some fixes and some comments

% Update the title and author information
\title{Learning to Estimate Search Progress Using Graph Neural Networks \thanks{This work expands on \citet{sudry2022learning}.}}
\author{
    Gur Keinan\textsuperscript{\rm 1},
    Naomi Derel\textsuperscript{\rm 1}
}
\affiliations{
    \textsuperscript{\rm 1}Technion - Israel Institute of Technology\\
    gur.keinan@campus.technion.ac.il, naomi.derel@campus.technion.ac.il
}

@@

Many real-world problems can be effectively modeled within classical search spaces and tackled using well-established heuristic search algorithms. The task of accurately predicting search progress is essential in several practical applications.
Beyond simply estimating completion time, accurate progress prediction allows better resource allocation, informs algorithm selection, and plays a vital role in temporal planning. In temporal planning scenarios \citep{cashmore2018temporal}, where actions have durations and deadlines, the ability to estimate remaining search time helps prioritize promising search paths that are more likely to yield timely solutions.

@@

% \naomi{cite or remove - Statistical approaches, such as those using pattern databases or sampling techniques, provide more flexibility but can be computationally intensive and may struggle with generalization across diverse problem domains.}

@@

Our work builds on this idea by integrating graph neural networks to capture and explicitly leverage the structure of the search space, learning over the graph formed by algorithms like A* \citep{hart1968formal}. This approach is novel in enabling the model to learn from both the features of individual nodes and the relationships between them. The underlying intuition is that this structural awareness supports more nuanced predictions by considering the broader context of the search process.

In this work, we aim to validate the effectiveness of our approach, investigating its capacity to enhance search progress predictions through structural insights. We developed a comprehensive framework for creating and analyzing search graphs of varying sizes and difficulties from classical planning problems, with a focus on 2 domains: Blocks-World and Sliding-Puzzle. These generated problem instances were solved using A* search with multiple domain-based and general heuristics, and the generated search trees were saved as graph structures. Our framework captures rich information about both the structure of the search graph, as well as rich node features such as heuristic values, depth in the search tree, branching factors, and various search progress indicators as introduced by \citet{sudry2022learning}.

We propose two primary GNN models: a heavier architecture and a more lightweight one. To assess their performance, we performed extensive experiments in various domains and at different search progress checkpoints. Our GNN models significantly outperformed previous approaches, reducing mean squared error (MSE) by an order of magnitude compared to both traditional formula-based estimators and a Random Forest (RF) model. More importantly, our models demonstrated strong generalization capabilities. When trained on one domain and tested on another, they maintained high performance with only a modest drop in accuracy—comparable to previous methods' performance when tested within the same training domain.

@@

In this section we briefly review the necessary background for heuristic search, our goal of search progress estimation, and the GNN architecture.

@@

A planning problem is formally represented as a state-space search problem, defined by a tuple $T = \langle S, S_0, S_g, A, f, c\rangle$, where: $S$ is the finite set of all possible states, $S_0$ is the initial state, $S_g$ is the set of goal states, $A(s)$ represents the set of actions applicable in state $s$, $f: S \times A \rightarrow S$ is the state transition function mapping a state and action to a resultant state, and $c(s,a)$ is the cost function for action $a$ performed from state $s$.

To solve planning problems efficiently, heuristic search algorithms like A* use a heuristic function $h(s)$ that estimates the cost from any state $s$ to a goal state. A* maintains an open list of nodes to explore, ordering them by $f(s) = g(s) + h(s)$, where $g(s)$ is the known cost from the initial state to $s$. The algorithm expands nodes by generating their successors until reaching a goal state or exhausting the search space.

@@

The search progress at any point is then defined as:
\begin{equation*}
    Prog_A(P) = \frac{Gen_A(P)}{Gen_A(P) + Rem_A(P, Gen_A(P))}
\end{equation*}

@@

\paragraph{Offline methods} attempt to predict search effort before the search begins by analyzing the search space structure.

\citet{breyer2008recent} focused on predicting A* node expansions for the 15-puzzle, developing analytical formulas based on pattern database statistics. 
\citet{lelis2014estimating} introduced a more general approach using sampling to estimate search tree size, where they performed random walks in the search space to estimate branching factors and solution depth. \citet{hutter2014algorithm} proposed methods for algorithm runtime prediction using instance features and machine learning, though this focused on overall runtime rather than progress during search. \citet{kilby2006estimating} showed that by weighting branches according to their probability of being visited during random probing, and by using recursive assumptions about search tree symmetry, search tree size could be estimated effectively without advance knowledge of the tree structure.

\paragraph{Online methods} make predictions during the search process using information from expanded nodes. \citet{thayer2012we} introduced several influential estimators:

@@

\paragraph{Learning to Estimate:}\label{para:learning-to-estimate}
The work of \citet{sudry2022learning} marked a significant shift by applying deep learning to this problem. Their LSTM-based approach processes sequences of expanded nodes, representing a significant shift towards applying deep learning to search progress estimation. For each node, they capture basic search node-features and global features:

@@

\paragraph{Graph Neural Networks in Planning:} GNNs have been successfully applied to various aspects of planning and search algorithms. \citet{toyer2020asnets} introduced Action Schema Networks (ASNets), which learn generalized policies for PDDL planning problems by exploiting their relational structure and can generalize from small training instances to larger problems. \citet{chen2024learning} developed graph representations for learning domain-independent heuristics, notably introducing a method that operates directly on lifted representations of planning tasks. \citet{staahlberg2022learning} employed a simplified GNN architecture for learning generalized policies in classical planning, establishing connections between GNN expressiveness and first-order logic fragments. \citet{9636442} demonstrated GNNs' utility in continuous search spaces by integrating them with Monte Carlo Tree Search for motion planning in autonomous driving. These diverse applications highlight GNNs' capability to effectively process and learn from structured planning and search spaces.

\paragraph{GNNs for Progress Estimation:} While GNNs have been widely applied to various graph-based tasks, their use for search progress estimation remains largely unexplored.
While LSTMs capture temporal patterns in the sequence of expanded nodes, they don't explicitly model the structural relationships in the search space. GNNs can directly operate on the graph structure of expanded nodes, where edges represent state transitions. This allows the model to learn from both node features and the topological structure of the search space, potentially enabling more accurate predictions by leveraging this additional structural information.

@@

The key to training effective GNN models for search progress estimation lies in having a diverse and representative dataset of search graphs. We created a comprehensive dataset by implementing and running A* search on two distinct planning domains, Blocks-World and Sliding-Puzzle, each with three different heuristic functions.

For the Blocks-World domain, we implemented two domain-specific heuristics. The misplaced blocks heuristic ($h_{\text{misplaced}}$) counts blocks not in their goal positions, providing a quick but non-admissible estimate - since each misplaced block may require multiple moves. The height difference heuristic ($h_{\text{height}}$) sums absolute differences between current and goal stack heights divided by 2, offering an admissible but sometimes uninformative estimate - when stack heights match but blocks are misplaced.

@@

We generated problem instances with varying parameters to ensure good coverage of the problem space. For Blocks World, we varied the number of blocks (5-10), number of stacks (3-5), and solution depth (7-15 moves). For Sliding Puzzles, we generated boards of different sizes (5x5 to 9x9) with varying solution depths (7-15 moves). For each parameter combination, we generated 200 random problem instances.

For each problem instance, we ran A* search with all three domain-specific heuristics, recording detailed information about the search process. Each node in the resulting search graphs contains:
\begin{itemize}
    \item Search featuresas described in \nameref{para:learning-to-estimate}.
    \item Search trajectory information (parent node, children count).
    \item The desired prediction - search progress ($\frac{\text{serial number}}{\text{total nodes}}$).
\end{itemize}

The raw dataset contains approximately 2.9GB of search graphs across both domains. Importantly, while these graphs originate from different domains and heuristics, they share a common structure and feature set that makes them domain-agnostic from the model's perspective. Each graph represents a general search trajectory with standardized node features and connectivity patterns, allowing us to treat them as a unified collection of search behaviors rather than domain-specific examples.

@@

To make the dataset more practical for GNN training and to better simulate real-world scenarios, we implemented a two-stage pruning strategy. First, we filtered out graphs with more than 15,000 nodes to manage computational complexity. Second, we applied a dynamic pruning approach where for each remaining graph, we randomly selected a threshold $t \in \{0.3, 0.5, 0.7\}$ and kept only nodes with $\frac{\text{serial\_number}}{\text{total\_nodes}} < t$.

@@

Each graph maintains parent-child relationships between nodes, representing the actual search tree structure explored by A*. When pruning nodes, we carefully preserve edge connections between remaining nodes to maintain the graph's structural integrity. This preservation of structural information, even in pruned graphs, is crucial for our GNN-based approach, differentiating it from previous sequence-based methods.

\paragraph{The final processed dataset} consists of:
\begin{itemize}
    \item Blocks-World: 7312 graphs averaging 2519 nodes per full search graph.
    \item Sliding-Puzzle: 4633 graphs averaging 495 nodes per full search graph.
\end{itemize}

@@

\subsection{Full-Scale Architecture}
The primary architecture, HeavyGNN, employs a multi-scale approach to capture both fine-grained node relationships and broader structural patterns in the search space. At its core, the model combines Graph Attention Networks (GAT) with Graph Convolutional Networks (GCN) in parallel processing streams. The GAT component uses multi-head attention mechanisms to identify important node relationships, while the GCN component captures neighborhood structures through weighted message passing. These parallel streams are combined through feature fusion, with layer normalization applied to maintain stable training dynamics.

@@

\subsection{Lightweight Architecture}
LightGNN provides an efficient alternative designed for scenarios where computational resources are limited or rapid inference is required. This model simplifies the architecture by using only GCN layers with improved neighborhood aggregation, removing the computationally expensive attention mechanisms while maintaining essential features for effective learning.

@@

@@

@@

    \item \textbf{Domain and Algorithm Coverage:} Our evaluation is currently limited to two classical planning domains (Blocks World and Sliding Puzzle) and a single search algorithm (A*). While these domains represent different types of search spaces, future work should evaluate and adapt the approach for a broader range of planning domains and search algorithms (e.g. GBFS, beam search). This would provide stronger validation of the method's generalizability and practical utility.

>>>>

\title{Learning to Estimate Search Progress Using Graph Neural Networks \thanks{This work expands on \citet{sudry2022learning}.}}
\author{
    Gur Keinan\textsuperscript{\rm 1},
    Naomi Derel\textsuperscript{\rm 1}
}
\affiliations{
    \textsuperscript{\rm 1}Technion - Israel Institute of Technology\\
    gur.keinan@campus.technion.ac.il, naomi.derel@campus.technion.ac.il
}

@@

Many real-world problems can be effectively modeled within classical search spaces and tackled using well-established heuristic search algorithms. When applying these algorithms, accurately predicting search progress is essential in several practical applications. Beyond simply estimating completion time, accurate progress prediction allows better resource allocation, informs algorithm selection, and plays a vital role in temporal planning scenarios \citep{cashmore2018temporal}. In these scenarios, where actions have durations and deadlines, the ability to estimate remaining search time helps prioritize promising search paths that are more likely to yield timely solutions.

@@

@@

In this work, we propose a novel approach using Graph Neural Networks (GNNs) to leverage this valuable structural information for more accurate search progress estimations. By learning on search graphs induced by informed search algorithmis like $A^{\star}$ \citep{hart1968formal}, where the nodes correspond to expanded states and edges to state transitions, the GNN can learn meaningful patterns in the search trajectory. This approach enables the model to learn from both the features of individual nodes and the relationships between them.

Throughout our project, we aim to validate the effectiveness of our approach, investigating its capacity to enhance search progress predictions through structural insights. We developed a comprehensive framework for creating and analyzing search graphs of varying sizes and difficulties from classical planning problems, with a focus on 2 domains: Blocks-World and Sliding-Puzzle. These generated problem instances were solved using $A^{\star}$ search with multiple domain-based and general heuristics, and the generated search trees were saved as graph structures. Our framework captures rich information about both the structure of the search graph, as well as rich node features such as heuristic values, depth in the search tree, branching factors, and various search progress indicators as introduced by \citet{sudry2022learning}.

We developed two distinct GNN models: a more complex architecture ("HeavyGNN") and a lightweight counterpart ("LightGNN"). To assess their performance, we performed extensive experiments in various domains and at different search progress checkpoints. Our GNN models significantly outperformed previous approaches, reducing mean squared error (MSE) by an order of magnitude compared to both traditional formula-based estimators and a Random Forest (RF) model. More importantly, our models demonstrated strong generalization capabilities. When trained on one domain and tested on another, they maintained high performance with only a modest drop in accuracy—comparable to previous methods' performance when tested within the same training domain.

@@

In this section, we briefly review the necessary background for the rest of the project, including search problems, heuristic search, Graph Neural Networks (GNNs), and the relevant literature on search progress estimation.

@@

A state-space search problem is defined by a tuple $T = \langle S, S_0, S_g, A, f, c\rangle$, where: $S$ is the finite set of all possible states, $S_0$ is the initial state, $S_g$ is the set of goal states, $A(s) \subset A$ represents the set of actions applicable in state $s$, $f: S \times A \rightarrow S$ is the state transition function mapping a state and action to a resultant state, and $c(s,a)$ is the cost function for action $a$ performed from state $s$.

To solve planning problems efficiently one might use informed search algorithms such as $A^{\star}$ \citep{hart1968formal}. Such algorithms use a heuristic function $h(s)$ that estimates the cost from any state $s$ to a goal state to guide exploration of the search space. For example, $A^{\star}$ maintains an open list of nodes to explore, ordering them by $f(s) = g(s) + h(s)$, where $g(s)$ is the known cost from the initial state to $s$. The algorithm expands nodes by generating their successors and appending them to the open list. The search is terminated when reaching a goal state or exhausting the search space.

@@

The search progress at any point is then defined as:

\begin{equation*}
    Prog_A(P) = \frac{Gen_A(P)}{Gen_A(P) + Rem_A(P, Gen_A(P))}
\end{equation*}

@@

\gur{The partition here is off, a lot of the 'offline' methods are online.}

\paragraph{Offline methods} Attempt to predict search effort before the search begins by analyzing the search space structure.

\citet{breyer2008recent} focused on predicting $A^{\star}$ node expansions for the 15-puzzle, developing analytical formulas based on pattern database statistics.
\citet{lelis2014estimating} introduced a more general approach using sampling to estimate search tree size, where they performed random walks in the search space to estimate branching factors and solution depth. \citet{hutter2014algorithm} proposed methods for algorithm runtime prediction using instance features and machine learning, though this focused on overall runtime rather than progress during search. \citet{kilby2006estimating} showed that by weighting branches according to their probability of being visited during random probing, and by using recursive assumptions about search tree symmetry, search tree size could be estimated effectively without advance knowledge of the tree structure.

\paragraph{Online methods} Make predictions during the search process using information from expanded nodes. \citet{thayer2012we} introduced several influential estimators:

@@

\paragraph{Learning to Estimate}\label{para:learning-to-estimate}
The work of \citet{sudry2022learning} marked a significant shift by applying deep learning to this problem. Their LSTM-based approach processes sequences of expanded nodes, representing a significant shift towards applying deep learning to search progress estimation. For each node, they capture basic search node-features and global features:

@@

\paragraph{Graph Neural Networks in Planning} GNNs have been successfully applied to various aspects of planning and search algorithms. \citet{toyer2020asnets} introduced Action Schema Networks (ASNets), which learn generalized policies for PDDL planning problems by exploiting their relational structure and can generalize from small training instances to larger problems. \citet{chen2024learning} developed graph representations for learning domain-independent heuristics, notably introducing a method that operates directly on lifted representations of planning tasks. \citet{staahlberg2022learning} employed a simplified GNN architecture for learning generalized policies in classical planning, establishing connections between GNN expressiveness and first-order logic fragments. \citet{9636442} demonstrated GNNs' utility in continuous search spaces by integrating them with Monte Carlo Tree Search for motion planning in autonomous driving. These diverse applications highlight GNNs' capability to effectively process and learn from structured planning and search spaces.

\paragraph{GNNs for Progress Estimation} While GNNs have been widely applied to various graph-based tasks, their potential for search progress estimation remains largely unexplored. Current approaches primarily rely on capturing temporal patterns in the sequence of expanded nodes but don't explicitly model the structural relationships in the search space. GNNs can directly operate on the graph structure of expanded nodes, where edges represent state transitions. This allows the model to learn from both node features and the topological structure of the search space, potentially enabling more accurate predictions by leveraging this additional structural information.

@@

The key to training effective GNN models for search progress estimation lies in having a diverse and representative dataset of search graphs. We created a comprehensive dataset by implementing and running $A^{\star}$ search on two distinct planning domains, Blocks-World and Sliding-Puzzle, each with three different heuristic functions.

For the Blocks-World domain, we implemented two domain-specific heuristics. The misplaced blocks heuristic ($h_{\text{misplaced}}$) counts blocks not in their goal positions. The height difference heuristic ($h_{\text{height}}$) sums absolute differences between current and goal stack heights divided by 2, offering an admissible but sometimes uninformative estimate - when stack heights match but blocks are misplaced.

@@

We generated problem instances with varying parameters to ensure good coverage of the problem space. For Blocks World, we varied the number of blocks \gur{(5-10), number of stacks (3-5), and solution depth (7-15 moves). For Sliding Puzzles, we generated boards of different sizes (5x5 to 9x9) with varying solution depths (7-15 moves)}. For each parameter combination, we generated 200 random problem instances.

For each problem instance, we ran $A^{\star}$ search with all three domain-specific heuristics, recording detailed information about the search process. Each node in the resulting search graphs contains:
\begin{itemize}
    \item Search featuresas described in \nameref{para:learning-to-estimate}.
    \item Search trajectory information (parent node, children count).
    \item The desired prediction - search progress ($\frac{\text{serial number}}{\text{total nodes}}$).
\end{itemize}

The raw dataset contains approximately 2.9GB of search graphs across both domains. There are 7312 graphs for Blocks-World, averaging 2519 nodes per graph, and 4633 graphs for Sliding-Puzzle, averaging 495 nodes per graph. Importantly, while these graphs originate from different domains and heuristics, they share a common structure and feature set that makes them domain-agnostic from the model's perspective. Each graph represents a general search trajectory with standardized node features and connectivity patterns, allowing us to treat them as a unified collection of search behaviors rather than domain-specific examples.

@@

To make the dataset more practical for GNN training and to better simulate real-world scenarios, we implemented a two-stage pruning strategy. First, we filtered out graphs with more than 15,000 nodes to manage computational complexity. Second, we applied a dynamic pruning approach where for each remaining graph, we randomly selected a threshold $t \in \{0.3, 0.5, 0.7\}$ and kept only nodes with $\frac{\text{serial number}}{\text{total nodes}} < t$.

@@

Each graph maintains parent-child relationships between nodes, representing the actual search tree structure explored by $A^{\star}$. When pruning nodes, we carefully preserve edge connections between remaining nodes to maintain the graph's structural integrity. This preservation of structural information, even in pruned graphs, is crucial for our GNN-based approach, differentiating it from previous sequence-based methods.

Notably, notice that unlike previous methods, in which a node only has information about nodes with smaller serial number, our model assumes that a parent node can have access to its children in inference. We don't find it problematic since any node in the search graph can equally contribute to estimating the remaining search effort - knowing the progress at any point can help predict how many more nodes need to be expanded to reach the goal. Therefore, if information about future nodes can help better estimate progress at an earlier node, there's no fundamental reason to restrict the model to only past knowledge. This view aligns with how search algorithms operate in practice, where both historical and frontier information contribute to understanding the search space.

@@

\subsection{Full-Scale Architecture}

\gur{GAT - global structure or only neighbors weighting?}

The primary architecture, HeavyGNN, employs a multi-scale approach to capture both fine-grained node relationships and broader structural patterns in the search space. At its core, the model combines Graph Attention Networks (GAT) with Graph Convolutional Networks (GCN) in parallel processing streams. The GAT component uses multi-head attention mechanisms to identify important node relationships, while the GCN component captures neighborhood structures through weighted message passing. These parallel streams are combined through feature fusion, with layer normalization applied to maintain stable training dynamics.

@@

\subsection{Lightweight Architecture}

LightGNN provides an efficient alternative designed for scenarios where computational resources are limited or rapid inference is required. This model simplifies the architecture by using only GCN layers with improved neighborhood aggregation, removing the computationally expensive attention mechanisms while maintaining essential features for effective learning.

@@

\gur{Change - test, all data.}

@@

\gur{Training details - train, eval, test.}

\gur{In the table report the error on the entire dataset, add a column of the test set error.}

@@

    \item \textbf{Domain and Algorithm Coverage:} Our evaluation is currently limited to two classical planning domains (Blocks World and Sliding Puzzle) and a single search algorithm ($A^{\star}$). While these domains represent different types of search spaces, future work should evaluate and adapt the approach for a broader range of planning domains and search algorithms (e.g. GBFS, beam search). This would provide stronger validation of the method's generalizability and practical utility.
